Turkey Hemorrhagic Enteritis Virus transcriptome profiling

### 2023-05-21 manuscript_thev_transcriptome.Rmd
Continue writing manuscript; write scripts to insert stats

```{r call_libs, echo=FALSE,include=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

@@

## ABSTRACT
#### Background:
Hemorrhagic enteritis (HE)is a disease affecting 6-12-week-old turkeys characterized by **immunosuppression (IS)**, bloody diarrhea, and up to 60% mortality. This disease is caused by **_Turkey Hemorrhagic Enteritis Virus (THEV)_** of which avirulent strains (THEV-A) that do not cause HE but retain the immunosuppressive ability have been isolated. The THEV-A Virginia Avirulent Strain (VAS) is still used as a live vaccine despite its immunosuppressive properties. **_Our objective is to understand the genetic basis by which VAS brings about IS_**. The **transcriptome of THEV** was studied to set the stage for further experimentation with specific viral genes that may mediate IS.

@@

#### Results:
Over 18 millionXXXXXX reads mapped to THEV genome providing good coverage/depth, leaving no regions unmapped. All predicted genes in the genome were represented. In keeping with all adenoviruses, all transcripts were spliced with either with 5'- or 3'-multi exon UTRs hitherto unknown. ThirteenXXXXX novel exons were identified which were validated by PCR and Sanger sequencing. The splicing patterns strongly suggest that there are 3XXXXX main promoters (E1, E3, and major late promoters) driving expression of most of the genes with 2XXXXXX possible minor promoters driving single genes (ORF7 and ORF8).

@@

Almost all AdV mRNAs are generated by the excision of one or more introns and most of these introns are located in the 5' or 3' UTRs of pre-mRNA. Thus the viral introns do only in a few cases interrupt the open reading frames (ORFs) [@Davison2003; @Zhao2014]. The development of high throughput sequencing methods has facilitated the discovery of many novel transcribed regions and splicing isoforms. It is also a very powerful tool to study alternative splicing under different conditions at an unparalleled depth [@Djebali2012; @Zhao2014]. Here, a paired-end deep sequencing experiment was performed to characterize for the first time, the transcriptome of THEV (VAS strain) during different phases of the infection, yielding a complete THEV splicing map. Our paired-end sequencing allowed for reading 150bp long sequences from each end of cDNA fragments. The data generated from paired-end sequencing, utilized in our experiment, should thus be reliable.
\newpage

## METHODS   
### Cell culture and THEV Infection   
The Turkey B-cell line (MDTC-RP19, ATCC CRL-8135) was grown as suspension cultures in 1:1 complete Leibovitz's L-15/McCoy's 5A medium with 10% fetal bovine serum (FBS), 20% chicken serum (ChS), 5% tryptose phosphate broth (TPB), and 1% antibiotics solution (100 U/mL Penicillin and 100ug/mL Streptomycin), at 41^o^C in a humidified atmosphere with 5% CO~2~. When infected, the cells were maintained in 1:1 serum-reduced Leibovitz's L15/McCoy's 5A media (SRLM) with 2.5% FBS, 5% ChS, 1.2% TPB, and  1% antibiotics solution (100 U/mL Penicillin and 100ug/mL Streptomycin). A commercially available HE vaccine was purchased from Hygieia Biological Labs as a source of THEV-A. The stock virus was titrated using an in-house qPCR assay with titer expressed as genome copy number(GCN)/mL. Cells were infected at a multiplicity of infection (MOI) of 100 GCN/cell and samples in triplicates were harvested at 4-, 12-, 24-, and 72-h.p.i for RNA sequencing. A second infection was done but samples in triplicates were harvested at 12-, 24-, 36-, 48-, and 72-h.p.i for PCR validation of novel splice sites.

@@

### Computational Analysis of RNA Sequencing Data: Mapping and Transcript characterization
Analysis of our sequencing reads were analyzed following a well established general outline described by Pertea _et al_ [@Pertea2016]. Briefly, sequencing reads were trimmed with the `FastQC`[@FastQC] program to acheive an overall `Mean Sequence Quality (Phred Score)` of 36. Trimmed reads were mapped to the complete sequence of avirulent turkey hemorrhagic enteritis virus strain Virginia (https://www.ncbi.nlm.nih.gov/nuccore/AY849321.1/) using `Hisat2` [@Pertea2016] with default settings except increasing the CPU cores to 10. The generated `BAM` files from each infection timepoint were fed into `StringTie` [@Pertea2016] using a `gff3` file from NCBI containing the predicted genes of THEV as a guide. A custom script was used to consolidate all transcripts from all timepoints without redundancy, generating the final transcriptome of THEV.

@@

## RESULTS

>>>>

```{r call_libs, echo=FALSE,include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

@@

```{r inline_values}

# extract paired-end sequencing read length
bam_info <- read_tsv("results/sgseq/bam_extracted_sampleData.tsv",
                     show_col_types = F) 
seq_len <- bam_info %>% select(read_length) %>% slice_head() %>% pull()

# extract coverage and mapping information
cov_all <- read_tsv("results/hisat2/coverage/bulk_coverage.txt",
                    comment = "Coverage", show_col_types = FALSE) %>%
  rename("organism" = "#rname") %>%
  filter(organism != "#rname") %>%
  mutate(timepoint = c("4hpi", "12hpi", "24hpi", "72hpi"),
         timepoint = factor(timepoint,
            levels = c("4hpi", "12hpi", "24hpi", "72hpi"))) %>%
  map_at(c(2:9), as.numeric) %>%
  as_tibble()

cov_4hr <- cov_all %>% filter(timepoint == "4hpi") %>%
              select(numreads) %>% pull()
cov_72hr <- cov_all %>% filter(timepoint == "72hpi") %>%
              select(numreads) %>% pull()
cov_72hr <- round(cov_72hr / 1e6, 2)
total_mapped <- round(sum(cov_all$numreads) / 1e6, 2)

load_count <- read_table("results/hisat2/coverage/bulk_counts.txt",
                         col_names = T, show_col_types = F) %>% pull()
tot_count <- round(mean(load_count) / 1e6, 2)
````
## ABSTRACT
#### Background:
Hemorrhagic enteritis (HE)is a disease affecting 6-12-week-old turkeys characterized by **immunosuppression (IS)**, bloody diarrhea, and up to 60% mortality. This disease is caused by **_Turkey Hemorrhagic Enteritis Virus (THEV)_** of which avirulent strains (THEV-A) that do not cause HE but retain the immunosuppressive ability have been isolated. The THEV-A Virginia Avirulent Strain (VAS) is still used as a live vaccine despite its immunosuppressive properties. **_Our objective is to understand the genetic basis by which VAS brings about IS_**. The **transcriptome of THEV** was studied to set the stage for further experimentation with specific viral genes that may mediate IS.

@@

#### Results:
A total of `r total_mapped` million reads mapped to THEV genome providing good coverage/depth, leaving no regions unmapped. All predicted genes in the genome were represented. In keeping with all adenoviruses, all transcripts were spliced with either with 5'- or 3'-multi exon UTRs hitherto unknown. ThirteenXXXXX novel exons were identified which were validated by PCR and Sanger sequencing. The splicing patterns strongly suggest that there are 3XXXXX main promoters (E1, E3, and major late promoters) driving expression of most of the genes with 2XXXXXX possible minor promoters driving single genes (ORF7 and ORF8).

@@

Almost all AdV mRNAs are generated by the excision of one or more introns and most of these introns are located in the 5' or 3' UTRs of pre-mRNA. Thus the viral introns do only in a few cases interrupt the open reading frames (ORFs) [@Davison2003; @Zhao2014]. The development of high throughput sequencing methods has facilitated the discovery of many novel transcribed regions and splicing isoforms. It is also a very powerful tool to study alternative splicing under different conditions at an unparalleled depth [@Djebali2012; @Zhao2014]. Here, a paired-end deep sequencing experiment was performed to characterize for the first time, the transcriptome of THEV (VAS strain) during different phases of the infection, yielding a complete THEV splicing map. Our paired-end sequencing allowed for reading `r seq_len` bp long sequences from each end of cDNA fragments. The data generated from paired-end sequencing, utilized in our experiment, should thus be reliable.
\newpage

## RESULTS      
**Overview of sequencing data and analysis pipeline outputs**     
A previous study by Zeinab *et al* showed that most THEV transcripts were detectable beginning at 4 hours[@Aboezz2019]; hence, infected MDTC-RP19 cells were collected in triplicates at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide window for transcript capture. Our `r seq_len`bp paired-end RNA sequencing experiment yielded an average of `r tot_count` million total reads per timepoint and the reads mapping to the virus genome increased dramatically from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i. Altogether, `r total_mapped` million reads from all timepoints mapped to THEV genome providing good coverage/depth, leaving no regions unmapped.

## METHODS   
### Cell culture and THEV Infection   
The Turkey B-cell line (MDTC-RP19, ATCC CRL-8135) was grown as suspension cultures in 1:1 complete Leibovitz's L-15/McCoy's 5A medium with 10% fetal bovine serum (FBS), 20% chicken serum (ChS), 5% tryptose phosphate broth (TPB), and 1% antibiotics solution (100 U/mL Penicillin and 100ug/mL Streptomycin), at 41^o^C in a humidified atmosphere with 5% CO~2~. When infected, the cells were maintained in 1:1 serum-reduced Leibovitz's L15/McCoy's 5A media (SRLM) with 2.5% FBS, 5% ChS, 1.2% TPB, and  1% antibiotics solution (100 U/mL Penicillin and 100ug/mL Streptomycin). A commercially available HE vaccine was purchased from Hygieia Biological Labs as a source of THEV-A (VAS strain). The stock virus was titrated using an in-house qPCR assay with titer expressed as genome copy number(GCN)/mL. Cells were infected at a multiplicity of infection (MOI) of 100 GCN/cell and samples in triplicates were harvested at 4-, 12-, 24-, and 72-h.p.i for RNA sequencing. A second infection was done but samples in triplicates were harvested at 12-, 24-, 36-, 48-, and 72-h.p.i for PCR validation of novel splice sites.

@@

### Computational Analysis of RNA Sequencing Data: Mapping and Transcript characterization
Analysis of our sequencing reads were analyzed following a well established general outline described by Pertea _et al_ [@Pertea2016]. Briefly, sequencing reads were trimmed with the `FastQC`[@FastQC] program to acheive an overall `Mean Sequence Quality (Phred Score)` of 36. Trimmed reads were mapped to the complete sequence of avirulent turkey hemorrhagic enteritis virus strain Virginia (https://www.ncbi.nlm.nih.gov/nuccore/AY849321.1/) using `Hisat2` (version `xxxxxx`) [@Pertea2016] with default settings except increasing the CPU cores to 10 without relying on known splice sites. The generated `BAM` files from each infection timepoint were fed into `StringTie` [@Pertea2016] using a `gff3` file from NCBI containing the predicted genes of THEV as a guide. A custom script was used to consolidate all transcripts from all timepoints without redundancy, generating the final transcriptome of THEV.

@@
